Type-theoretical Semantics of \Pio\ via the Delay Monad \thanks{Supported by organization x.}

### 2019-04-25 mpc19/paper.tex
working on dagger trace

% Delay macros
\newcommand{\Delay}{\mathsf{Delay}\,}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\laterR}{\mathsf{laterR}}
\newcommand{\laterL}{\mathsf{laterL}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\costr}{\mathsf{costr}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}
\newcommand{\copairD}[2]{[#1,#2]_{\D}}
\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
\newcommand{\inlD}{\mathsf{inl}_{\D}}
\newcommand{\inrD}{\mathsf{inr}_{\D}}
\newcommand{\fstD}{\mathsf{fst}_{\D}}
\newcommand{\sndD}{\mathsf{snd}_{\D}}
\newcommand{\piso}{\mathsf{isPartialIso}}
\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
\newcommand{\iterD}{\mathsf{iter}_\D}
\newcommand{\unfolding}{\mathsf{unfolding}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\traceD}{\mathsf{trace}_\D}
\newcommand{\traceHD}{\mathsf{traceH}_\D}

@@

\section{TODO} % (fold)
\label{sec:todo}
Robin:
\begin{itemize}
  \item Abstract.
  \item Introduction.
  \item Make pictures! 
  \begin{itemize}
    \item Graphical language of monoidal categories, and for \Pio.
    \item Trace.  
    \item Iterator, iterator unfolding axiom.
    \item Trace from iterator.
  \end{itemize}
  \item Conclusion
  \begin{itemize}
    \item Related work: Kleisli category of the Delay monad is a join
    restriction category, inverse category of a join restriction category is a
    join inverse category; already known to be dagger-traced under certain
    conditions.
    \item Future work: Semantics of time invertible programming. Extension of
    the work on the Kleisli category of the Delay monad beyond $\Set$ to other
    (more general) categories, e.g. topoi.
  \end{itemize}
\end{itemize}
Niccolò: 
\begin{itemize}
  \item Fix the syntax to be closer to ordinary presentations (but do describe 
  how it is realized in Agda).
  \item Finish the section about trace and iteration.
  \item Remind the reader about double lines of coinductive definitions. 
\end{itemize}
Both:
\begin{itemize}
  \item \textbf{Important:} Come up with a punnier title! (Something about 
  delays and daggers).
\end{itemize}
% section todo (end)

@@

In order to model the reversible programs of \Pio, we need to consider
reversible computations in $\Dapprox$. Given a morphism $f : A \to
\Delay B$, we say that it is a partial isomorphism if the following
type is inhabited:
\begin{align*}
\piso \,f = (g : B \to \Delay A) \times \left( (a : A) (b : B) \to  f
  \,a \dn b \leftrightarrow g \,b \dn a  \right)  
\end{align*}
In other words, $f$ is a partial isomorphism if there exists a
morphism $g : B \to \Delay A$ such that, if $f\,a$ terminates
returning a value $b$, then $g\,b$ terminates returning $a$, and vice
versa. In $\Dapprox$, our definition of partial isomorphisms is equivalent to
the standard categorical one~\cite{Kastl79} (see also \cite{CockettL02}),
which, translated in our type-theoretical setting, is
\begin{align*}
\pisoalt \,f = (g : B \to \Delay A) \times f \diamond g \diamond
  f \approx f \times g \diamond f \diamond g \approx g 
\end{align*}
We denote $A \simeq B$ the type of partial isomorphisms between
$A$ and $B$:
\[
A \simeq B = (f : A \to \Delay B) \times \piso \,f
\]
We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting of
(equivalence classes of) partial isomorphisms. Note that $\Inv\Dapprox$
inherits neither partial products nor coproducts of $\Dapprox$, as the
universal mapping property fails in both cases. However, it can be shown that
in the category $\Inv\Dapprox$, $0$ is a zero object, $A + B$ is the
disjointness tensor product of $A$ and $B$ (in the sense of Giles~\cite{Giles})
with unit $0$, and $A \times B$ a monoidal product of $A$ and $B$ with unit $1$
(though it is \emph{not} an inverse product in the sense of
Giles~\cite{Giles}, as that would imply decidable equality on all objects).

@@

An Elgot monad is a monad $T$ whose Kleilsi category supports
unguarded Elgot iteration. More precisely\footnote{Here we give the
  definition of Elgot monad on $\Set$, but the definition of Elgot
  monad makes sense in any category with binary coproducts.}, a monad
$\T$ is Elgot if there exists an operation
\[
\iter_\T : (A \to \T \,(A + B)) \to A \to \T\,B
\]
satisfying a number of axioms, most notably the unfolding axiom:
\[
\iter_\T\,f \equiv \copair{\iter_\T \,f}{\eta_\T} \diamond_\T f
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.

The delay monad is an Elgot monad for which the axioms holds up to
weak bisimilarity, not propositional equality. Its iteration operator
relies on the definition of an auxiliary function $\iterD'$
corecursively defined as follows:
\[ %\arraycolsep=8pt
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B} \\
\iterD' \,f \,(\now \,(\inl\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
\iterD' \,f \,(\now \,(\inr\, b)) &= \now \,b \\
\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
 \\
\multicolumn{2}{l}{\iterD : (A \to \Delay (A + B)) \to A \to \Delay B} \\
\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
\end{array}
\]

%% \begin{align*}
%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
%% & \\
%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
%% \end{align*}
The definition above can be given the following intuitive
explaination. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
not terminate as well. If $f\,a$ terminates, there are two
possibilities: either $f\,a$ converges to $\inr\,b$, in which case
$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
to $\inl\,a'$, in which case we repeat the precedure by replacing $a$
with $a'$. Notice that in the latter case we also add one occurrence
of $\later$ to the total computation time. This addition is necessary
achieving the productivity of the corecursively defined function
$\iterD'$. In fact, by changing the first line of its specification to
$\iterD' \,f \,(\now \,(\inl\, a)) = \iterD' \,f \,(f a)$
and taking $f = \inlD$, we would have that $\iterD' \,f \,(\now
\,(\inl\, a))$ unfolds indefinitely without producing any output. In
Agda, such a definition would be rightfully rejected by the termination
checker.

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the unfolding
axiom. We employ the definition (\ref{eq:weak2}) of weak
bisimilarity. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (A + B)) \to
  \bind\, \copairD{\iterD \,f}{\now} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\now \,(\inr \,b)) & = \now_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (A + B)) \to \copairD{\iterD
  \,f}{\now} \diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

From the Elgot iteration operator it is possible to derive a trace
operator. First, given $f : A + B \to C$, we introduce
$\mapL f = f \comp \inl : A \to C$ and
$\mapR f = f \comp \inr : B \to C$, so that $f = [\mapL f , \mapR f]$.
The trace operator in $\Dapprox$ is defined as follows:
\begin{align*}
& \traceD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
& \traceD \,f = \copairD{\iterD \,\mapL f}{\now} \diamond \mapR f
\end{align*}
The operation $\traceD$ is compatible with weak bisimilarity.

\ctikzfig{fL_fR}
\ctikzfig{trace}

It is well-known that a trace operator is obtainable from an iteration
operator, as shown by Hasegawa  \cite{Hasegawa97}. His construction,
instantiated to our setting, looks like this:
\begin{align*}
& \traceHD : (A + B \to \Delay (A + C)) \to B \to \Delay C \\
& \traceHD \,f = \iterD (\Delay (\inl + \id) \comp f) \comp \inr
\end{align*}

@@

>>>>

% Delay macros
\newcommand{\Delay}{\mathsf{Delay}\,}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\laterR}{\mathsf{laterR}}
\newcommand{\laterL}{\mathsf{laterL}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\costr}{\mathsf{costr}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}
\newcommand{\copairD}[2]{[#1,#2]_{\D}}
\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
\newcommand{\inlD}{\mathsf{inl}_{\D}}
\newcommand{\inrD}{\mathsf{inr}_{\D}}
\newcommand{\fstD}{\mathsf{fst}_{\D}}
\newcommand{\sndD}{\mathsf{snd}_{\D}}
\newcommand{\piso}{\mathsf{isPartialIso}}
\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
\newcommand{\iterD}{\mathsf{iter}_\D}
\newcommand{\unfolding}{\mathsf{unfolding}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\traceD}{\mathsf{trace}_\D}
\newcommand{\traceHD}{\mathsf{traceH}_\D}
\newcommand{\daggerD}{\mathsf{dagger}_\D}
\newcommand{\Orb}[3]{\mathsf{Orb}\,#1\,#2\,#3}
\newcommand{\done}{\mathsf{done}}
\renewcommand{\next}{\mathsf{next}}
\newcommand{\reverseOrbit}{\mathsf{reverseOrb}}

@@

\section{TODO} % (fold)
\label{sec:todo}
Robin:
\begin{itemize}
  \item Abstract.
  \item Introduction.
  \item Make pictures! 
  \begin{itemize}
    \item Hasegawa trace from iterator.
  \end{itemize}
  \item Conclusion
  \begin{itemize}
    \item Related work: Kleisli category of the Delay monad is a join
    restriction category, inverse category of a join restriction category is a
    join inverse category; already known to be dagger-traced under certain
    conditions.
    \item Future work: Semantics of time invertible programming. Extension of
    the work on the Kleisli category of the Delay monad beyond $\Set$ to other
    (more general) categories, e.g. topoi.
  \end{itemize}
\end{itemize}
Niccolò: 
\begin{itemize}
  \item Fix the syntax to be closer to ordinary presentations (but do describe 
  how it is realized in Agda).
  \item Finish the section about trace and iteration.
  \item Remind the reader about double lines of coinductive definitions.
  \item Swap types in Elgot iterator.
\end{itemize}
Both:
\begin{itemize}
  \item \textbf{Important:} Come up with a punnier title! (Something about 
  delays and daggers).
\end{itemize}
% section todo (end)

@@

In order to model the reversible programs of \Pio, we need to consider
reversible computations in $\Dapprox$. Given a morphism $f : A \to
\Delay B$, we say that it is a partial isomorphism if the following
type is inhabited:
\begin{align*}
\piso \,f = (g : B \to \Delay A) \times \left( (a : A) (b : B) \to  f
  \,a \dn b \leftrightarrow g \,b \dn a  \right)  
\end{align*}
In other words, $f$ is a partial isomorphism if there exists a
morphism $g : B \to \Delay A$ such that, if $f\,a$ terminates
returning a value $b$, then $g\,b$ terminates returning $a$, and vice
versa. Given a partial isomorphism $f$, we denote its partial inverse
by $\daggerD\,f$.

In $\Dapprox$, our definition of partial isomorphisms is equivalent to
the standard categorical one~\cite{Kastl79} (see also \cite{CockettL02}),
which, translated in our type-theoretical setting, is
\begin{align*}
\pisoalt \,f = (g : B \to \Delay A) \times f \diamond g \diamond
  f \approx f \times g \diamond f \diamond g \approx g 
\end{align*}
We denote $A \simeq B$ the type of partial isomorphisms between
$A$ and $B$:
\[
A \simeq B = (f : A \to \Delay B) \times \piso \,f
\]
We call $\Inv \Dapprox$ the subcategory of $\Dapprox$ consisting of
(equivalence classes of) partial isomorphisms. Note that $\Inv\Dapprox$
inherits neither partial products nor coproducts of $\Dapprox$, as the
universal mapping property fails in both cases. However, it can be shown that
in the category $\Inv\Dapprox$, $0$ is a zero object, $A + B$ is the
disjointness tensor product of $A$ and $B$ (in the sense of Giles~\cite{Giles})
with unit $0$, and $A \times B$ a monoidal product of $A$ and $B$ with unit $1$
(though it is \emph{not} an inverse product in the sense of
Giles~\cite{Giles}, as that would imply decidable equality on all objects).

@@

An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleilsi category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of Elgot monad on
  $\Set$, but the definition of Elgot monad makes sense in any
  category with binary coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying a number of axioms, most notably the unfolding axiom:
\[
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.

The delay monad is an Elgot monad for which the axioms holds up to
weak bisimilarity, not propositional equality. Its iteration operator
relies on the definition of an auxiliary function $\iterD'$
corecursively defined as follows:
\[ %\arraycolsep=8pt
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\iterD' : (A \to \Delay (B + A)) \to \Delay (B + A) \to \Delay B} \\
\iterD' \,f \,(\now \,(\inl\, b)) &= \now \,b \\
\iterD' \,f \,(\now \,(\inr\, a)) &= \later \,(\iterD' \,f \,(f \,a)) \\
\iterD' \,f \,(\later \,x) &= \later \,(\iterD' \,f \,x) \\
 \\
\multicolumn{2}{l}{\iterD : (A \to \Delay (B + A)) \to A \to \Delay B} \\
\multicolumn{2}{l}{\iterD \,f \,a = \iterD' \,f \,(f \,a)}
\end{array}
\]

%% \begin{align*}
%% & \iterD' : (A \to \Delay (A + B)) \to \Delay (A + B) \to \Delay B \\
%% & \iterD' \,f \,(\now \,(\inl\, a)) = \later \,(\iterD' \,f \,(f \,a)) \\
%% & \iterD' \,f \,(\now \,(\inr\, b)) = \now \,b \\
%% & \iterD' \,f \,(\later \,x) = \later \,(\iterD' \,f \,x) \\ 
%% & \\
%% & \iterD : (A \to \Delay (A + B)) \to A \to \Delay B \\
%% & \iterD \,f \,a = \iterD' \,f \,(f \,a)
%% \end{align*}
The definition above can be given the following intuitive
explaination. If $f\,a$ does not terminate, then $\iterD\,f\,a$ does
not terminate as well. If $f\,a$ terminates, there are two
possibilities: either $f\,a$ converges to $\inr\,b$, in which case
$\iterD\,f\,a$ terminates returning the value $b$; or $f\,a$ converges
to $\inl\,a'$, in which case we repeat the precedure by replacing $a$
with $a'$. Notice that in the latter case we also add one occurrence
of $\later$ to the total computation time. This addition is necessary
for ensuring the productivity of the corecursively defined function
$\iterD'$. In fact, by changing the second line of its specification to
$\iterD' \,f \,(\now \,(\inr\, a)) = \iterD' \,f \,(f a)$
and taking $f = \inrD$, we would have that $\iterD' \,f \,(\now
\,(\inr\, a))$ unfolds indefinitely without producing any output. In
Agda, such a definition would be rightfully rejected by the termination
checker.

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the unfolding
axiom. We employ the definition (\ref{eq:weak2}) of weak
bisimilarity. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

From the Elgot iteration operator it is possible to derive a trace
operator. First, given $f : A + B \to C$, we introduce
$\mapL f = f \comp \inl : A \to C$ and
$\mapR f = f \comp \inr : B \to C$, so that $f = [\mapL f , \mapR f]$.
Graphically:
\ctikzfig{fL_fR}

The trace operator in $\Dapprox$ is defined as follows:
\begin{align*}
& \traceD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
& \traceD \,f = \copairD{\now}{\iterD \,\mapR f}\diamond \mapL f
\end{align*}
The operation $\traceD$ is compatible with weak bisimilarity. 
Pictorially:

\ctikzfig{trace}

Intuitively, the function $\mapL f$ initialises the loop. It either
diverges, so that the trace of $f$ diverges as well, or it terminates. It either terminates with
an element $b : B$, in which case the loop ends immediately returning
$b$, or it converges to a value $c : C$, and in this case we proceed
by  invoking the iteration of $\mapR f$ on $c$.

It is well-known that a trace operator is obtainable from an iteration
operator, as shown by Hasegawa  \cite{Hasegawa97}. His construction,
instantiated to our setting, looks as follows:
\begin{align*}
& \traceHD : (A + C \to \Delay (B + C)) \to A \to \Delay B \\
& \traceHD \,f = \iterD (\Delay (\id + \inr) \comp f) \comp \inl
\end{align*}

It is not difficoult to prove that the two possible ways of defining a
trace operator from Elgot iteration are equivalent, in the sense that
$\traceD\,f \approx \traceHD\,f$ for all $f : A + C \to \Delay (B +
C)$.

The trace axioms follow from the Elgot iteration axioms.

@@

We now show that $\traceD$ is a $\dagger$-trace operator,
i.e. if $f$ is a partial isomorphism, then
$\traceD\,f$ is also a partial isomorphism with partial inverse $\traceD\,(\daggerD\,f)$.

This is proved by introducing the notion of \emph{orbit} of an element
$x : A + C$ wrt. a function $f : A + C \to \Delay (B + C)$. The orbit of
$x$ consists of the terms of type $B + C$ that are obtained in a
finite number of steps from repeated applications of the function $f$
on $x$. Formally, a term $y$ belongs to the orbit of $f$ wrt. $x$ if
the type $\Orb f x y $ is inhabited, with the latter type
inductively defined as: 
\[
\infer{\done\,p : \Orb f x y}{p : f \,x \dn y}
\qquad 
\infer{\next\,p \,q : \Orb f x y}{p : f \,x \dn \inr\,c & q : \Orb f
  {(\inr\,c)} y}
\]

The notion of orbit can be used to state when 
the iteration of a function $f : A \to \Delay (B + A)$ on a input
$a : A$ terminates with value $b : B$.
\[
\iterD\,f\,a \dn b \leftrightarrow \Orb {\copairD{\inlD}{f}} {(\inr\,a)} {(\inl\,b)}
\]
We refer the intersted reader to our Agda formalization for a complete
proof of this logical equivalence.
Similarly, the orbit can be used to state when 
the trace of a function $f : A + C \to \Delay (B + C)$ on a input
$a : A$ terminates with value $b : B$.
\begin{equation}\label{eq:traceorbit}
\traceD\,f\,a \dn b \leftrightarrow \Orb f{(\inl\,a)} {(\inl\,b)}
\end{equation}

Showing that $\traceD$ is a $\dagger$-trace operator requires the
construction of an inhabitant of $\traceD\,f\,a \dn b \leftrightarrow
\traceD\,(\daggerD\,f)\,b \dn a$. Thanks to the logical equivalence in
(\ref{eq:traceorbit}), this is equivalent to prove the following
statement instead: 
\[
\Orb f {(\inl\,a)} {(\inl\,b)} \leftrightarrow \Orb {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}
\]
We give a proof of the left-to-right direction, the other implication
is proved in an analogous way.  Notice that a term
$p : \Orb f {(\inl\,a)} {(\inl\,b)}$ can be seen as a finite sequence of
elements of $C$, precisely the intermediate values produced by
$\traceD\,f\,a$ before converging to $b$. The orbit of $b$ wrt. the
partial inverse of $f$ can therefore be computed by reversing the
sequence of elements present in $p$. The construction of the reverse
of an orbit is very similar to the way the reverse of a list is typically
defined in a functional programming language like Haskell.

\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\reverseOrbit' : (i : \{a : A\} \{b : B\} \to f \,a
\dn b \to \daggerD\,f  \,b \dn a) \to } \\
\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \{c : C\} \to } \\
\multicolumn{2}{l}{\qquad \Orb f {(\inr\,c)}
 {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inr\,c)} {(\inl\,a)} \to } \\
\multicolumn{2}{l}{\qquad \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
\reverseOrbit' \,i\, (\done \,p) \,p'&= \next\, (i \,p)\,p' \\
\reverseOrbit' \,i\,(\next\,p\,q) \,p'&= \reverseOrbit' \,i\,q \,(\next \,(i\,p)\,p')
\end{array}
\]

\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\reverseOrbit : (i : \{a : A\} \{b : B\} \to f \,a
  \dn b \to \daggerD\,f  \,b \dn a) \to} \\
\multicolumn{2}{l}{\qquad \{a : A\} \{b : B\} \to} \\
\multicolumn{2}{l}{\qquad \Orb f {(\inl\,a)}
  {(\inl\,b)} \to \Orb  {(\daggerD\,f)} {(\inl\,b)} {(\inl\,a)}}\\
\reverseOrbit \,i\, (\done \,p) &= \done \, (i \,p) \\
\reverseOrbit \,i\,(\next\,p\,q) &= \reverseOrbit' \,i\,q \,(\done \,(i\,p))
\end{array}
\]
